---
pagetitle: GWAS
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE, purl=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
knitr::opts_knit$set(root.dir = "../course_files")
set.seed(123)
```

# Statistical models for GWAS

:::{.callout-tip}
#### Learning objectives

- Summarise the statistical methods used by PLINK to carry out trait-genotype association tests.
- Interpret the effect sizes of an association in relation to individual genotypes. 
- Distinguish how the interpretation of effect sizes differs for quantitative and binary trait models.
:::


## Association tests

At its simplest, the association test carried out by software such as PLINK is based on linear regression models. 
Most GWAS software support two types of regression models, depending on the nature of the trait: **normal** regression for quantitative traits or **binomial (logistic)** regression for binary traits (case/control, yes/no, presence/absence).

We go over the intuition for each of these approaches.
In the examples below we assume there are two alleles at a variant (i.e. it is bi-allelic), with alleles "A" and "B". 
The three diploid genotypes are written as "AA", "AB" and "BB".


## Quantitative trait models

For **quantitative traits** we assume that the trait is well-approximated by a Normal distribution, $Trait \sim N(\mu, \sigma^2)$ with mean $\mu$ and variance $\sigma^2$. 
The mean of this distribution is modelled as dependent on genotype, and we test for the significance of this relationship.

```{r regression-sim}
#| echo: false
library(tidyverse) # data manipulation
library(patchwork) # to compose plots
library(janitor)   # to clean column names
theme_set(theme_minimal()) # change default ggplot2 theme
set.seed(42)

xs <- sample(0:2, 100, replace = TRUE, prob = c(0.25, 0.5, 0.25))
ys <- rnorm(length(xs), model.matrix(~ xs) %*% c(0, 1))

tibble(genotypes = xs, trait = ys) |> 
  ggplot(aes(factor(genotypes), trait)) +
  geom_jitter(width = 0.1) +
  geom_smooth(aes(group = 1), 
              method = "lm", 
              se = FALSE, 
              colour = "firebrick", 
              size = 2) +
  labs(x = "Genotype (number of 'B' alleles)", y = "Trait") +
  theme_classic() +
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank())
```

The strength of the relationship between genotype and trait is measured by the slope of the regression line, often referred to as **Beta (β)**. 

For example, if our trait was blood pressure measured in _mm Hg_, and we estimated β = 3 for a given variant, this would be the interpretation of the results:

- Genotype AA: baseline group (carrying zero "B" alleles).
- Genotype AB: +3 mg Hg compared to AA.
- Genotype BB: +6 mg Hg compared to AA.

This is a so-called "additive genetic model", where we code the genotypes as "number of B alleles" and we thus assume that homozygotes "BB" increase by twice relative to heterozygotes. 

The **significance of association** is determined by testing for the slope of the regression being zero (i.e. a flat line). 
This is done by using a t-statistic, which measures the "signal-to-noise" ratio of our estimate, and comparing it to a theoretical distribution to obtain a **p-value** for the association. 

:::{.callout-note}
#### Quantitative traits

Quantitative traits may include both continuous type of data (e.g. bmi, height) as well as and discrete count data (e.g. cell count, cigarette consumption). 
In generalised linear models, count data would usually be modelled using a more appropriate distribution rather than the Normal (e.g. a Poisson or Negative Binomial regression). 

However, these methods can be computationally expensive and so, in practice, software packages for GWAS usually only implement the normal regression model for quantitative traits. 
:::


## Binary trait models

Conceptually, things are very similar for **binary traits**, except we assume the trait comes from a binomial (think of coin flips 🪙): $Trait \sim Binomial(n, p)$ with known sample size $n$ (that's how many individuals we have) and a probability $p$ that the event happens. 
The probability $p$ is modelled as depending on genotype, and we test for the significance of this relationship. 

```{r logistic-sim}
#| echo: false
set.seed(42)

xs <- sample(0:2, 100, replace = TRUE, prob = c(0.25, 0.5, 0.25))
# I define the betas relative to their original scale
# the intercept is the log(odds) for genotype 0, which I set by defining the baseline probability as 10%
# the beta is the log(odds) for each allele, which I set to 7
ys <- rbinom(length(xs), 
             size = 1, 
             prob = plogis(model.matrix(~ xs) %*% c(log(0.1/(1-0.1)), log(3))))

tibble(genotypes = xs, trait = ys) |> 
  ggplot(aes(genotypes, trait)) +
  geom_jitter(width = 0.1, height = 0.05) +
  geom_smooth(aes(group = 1), 
              method = "glm", 
              method.args = list(family = "binomial"),
              se = FALSE, 
              colour = "firebrick", 
              size = 2) +
  labs(x = "Genotype (number of alternative allele copies)", 
       y = "Trait probability") +
  theme_classic() +
  scale_x_continuous(breaks = 0:2)
```

For binary traits, the strength of the relationship between trait and genotype is expressed as an **odds ratio (OR)**, i.e. the ratio of the probability that the event occurs over the probability that it doesn't: 

$$
\text{Odds Ratio} = \frac{p}{1 - p}
$$

Taking a concrete example, if our trait is "type 2 diabetes", the odds ratio represents the probability change in having this condition, relative to not having it:

$$
\text{Odds Ratio} = \frac{\text{probability diabetes}}{\text{probability healthy}}
$$

One important thing to consider is that the coefficients of logistic regression models are multiplicative.
For example, if the estimate for a given variant was OR = 3, the interpretation in terms of genotypes is:

- Genotype AA: baseline group (carrying zero "B" alleles).
- Genotype AB: 3 times more likely to have the condition than AA.
- Genotype BB: 3² = 9 times more likely to have the condition than AA.

The **significance of association** in this case is assessed using a Wald Z-statistic, which similarly to the t-statistic discussed earlier, measures the "signal-to-noise" ratio of our OR estimate.
As usual, this is compared to a theoretical distribution to obtain a **p-value** for the association.


:::{.callout-important}
#### Effect sizes

In the examples given above we used relatively large effect sizes, for illustration purposes. 
For complex traits the effects of individual variants might be quite small. 
For example, for type 2 diabetes many found associations have OR of ~1.1 or even less. 
For reference, this represents a probability of 52% of having the disease compared to 48% for not having it. 

These are still substantial effects, but not as large as what we used in the examples above. 
:::


:::{.callout-note}
#### Odds, log-odds and probabilities

Logistic regression models can be challenging to interpret and working with odds is sometimes confusing. 
Remember, the odds is a ratio of two probabilities: the probability of an event happening versus it not happening. 

Because of the underlying maths, logistic regression coefficients are usually given as log-odds. 
This is exactly how it sounds like, it's the natural log of the odds ratio. 
This is also known as the logit function.

I can be useful to convert probabilities to odds (or log-odds) and back. 
Say an event has a 75% chancde of happening, we can calculate odds and log-odds easily: 

```{r}
# if you know your probability
p <- 0.75

# calculate odds as:
0.75 / (1 - 0.75)

# and log-odds taking the natural log of the previous
log(0.75 / (1 - 0.75))
```

On the other hand, if you start from having the odds ratio, for example OR = 3, then you need to: 
first convert it to log-odds (or "logit"); then use the inverse-logit function to calculate the probability.
Here's the R code: 

```{r}
# if you know your odds
odds <- 3

# calculate the probability of the event using the inverse-logit
plogis(log(odds))
```

:::


## Summary

::: {.callout-tip}
#### Key Points

- TODO
:::
