---
pagetitle: GWAS
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE, purl=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
knitr::opts_knit$set(root.dir = "../course_files")
set.seed(123)
```

# Visualising GWAS

:::{.callout-tip}
#### Learning objectives

- Produce Manhattan plots to visualise the association results across the genome.
- Explore the association results and refine visualisations to focus on particular regions of interest.
:::


## Manhattan plot

```{r}
# load the libraries
library(tidyverse) # data manipulation
library(patchwork) # to compose plots
library(janitor)   # to clean column names
theme_set(theme_minimal())
```


Import the data and create custom theme for Manhattan plots: 

```{r}
coffee <- read_tsv("results/1000G_subset_nocovar.coffee.glm.linear") |> 
  clean_names()
  
manhattan_theme <- theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid = element_blank(),
    panel.spacing = unit(0.1, "lines"),
    strip.background = element_blank()
  )
```

We can import and visualise the results:

```{r}
coffee_pca <- read_tsv("results/1000G_subset_pca.coffee.glm.linear") |> 
  clean_names()
  
# retain only the SNP test results
coffee_pca <- coffee_pca |> 
  filter(test == "ADD")

# histogram of p-values
coffee_pval_hist <- coffee_pca |> 
  arrange(p) |> 
  mutate(expected = -log10(ppoints(n())), 
         observed = -log10(p)) |> 
  filter(p <= 0.001 | (p > 0.001 & runif(n()) < 0.01)) |> 
  ggplot(aes(p)) +
  geom_histogram(binwidth = 0.01) +
  labs(title = "P-value histogram")

# qqplot
coffee_pval_qq <- coffee_pca |> 
  arrange(p) |> 
  mutate(expected = -log10(ppoints(n())), 
         observed = -log10(p)) |> 
  filter(p <= 0.001 | (p > 0.001 & runif(n()) < 0.01)) |> 
  ggplot(aes(expected, observed)) +
  geom_point() +
  geom_abline() + 
  labs(title = "Q-Q plot")

# manhattan plot
coffee_pval_man <- coffee_pca |> 
  filter(p < 0.01) |> 
  ggplot(aes(pos, -log10(p))) +
  geom_point() +
  geom_hline(yintercept = -log10(5e-8), linetype = "dashed") +
  facet_grid(~ number_chrom, 
             scale = "free_x", 
             space = "free_x",
             switch = "x") +
  labs(x = "Chromosome", 
       title = "Manhattan plot") +
  manhattan_theme
```

Compose plots: 

```{r}
((coffee_pval_hist + coffee_pval_qq) / coffee_pval_man) + 
  plot_annotation(
    title = "Caffeine consumption (mg/day)"
  )
```

Inflation factor:

```{r}
median(qchisq(coffee_pca$p, df=1, lower.tail = F), na.rm = T)/qchisq(0.5, 1)
```


## Regional plots

```{r}
# Zoom in on one of the SNPs
coffee_pca |> 
  filter(number_chrom == 7 & pos > 17244953 - 250e3 & pos < 17244953 + 250e3) |> 
  mutate(label = ifelse(p == min(p), 
                        paste(number_chrom, pos, sep = ":"),
                        NA)) |> 
  ggplot(aes(pos, -log10(p))) +
  geom_point() +
  geom_text(aes(label = label), hjust = -0.1) +
  geom_hline(yintercept = -log10(5e-8), linetype = "dashed")
```

Calculate LD for target SNP:

```bash
plink2 --pfile data/plink/1000G_subset --out results/1000G_subset \
  --geno 0.05 --maf 0.01 --hwe 0.001 keep-fewhet \
  --mind 0.05 --keep results/1000G_subset.king.cutoff.in.id \
  --r2-unphased --ld-window-kb 500 --ld-snp rs4410790
```

Visualise:

```{r}
hit <- read_tsv("results/1000G_subset.vcor") |> 
  clean_names()

coffee_pca |> 
  filter(number_chrom == 7 & pos > 17244953 - 250e3 & pos < 17244953 + 250e3) |> 
  mutate(label = ifelse(p == min(p), 
                        paste(number_chrom, pos, sep = ":"),
                        NA)) |> 
  left_join(hit, by = c("id" = "id_b")) |> 
  mutate(unphased_r2 = ifelse(is.na(unphased_r2), 0, unphased_r2)) |> 
  mutate(unphased_r2 = ifelse(p == min(p), 1, unphased_r2)) |> 
  ggplot(aes(pos, -log10(p))) +
  geom_point(aes(colour = unphased_r2, size = unphased_r2)) +
  geom_text(aes(label = label), hjust = -0.1) +
  geom_hline(yintercept = -log10(5e-8), linetype = "dashed") +
  scale_colour_gradient2(low = "#313695", 
                         mid = "#ffffbf", 
                         high = "#a50026", 
                         midpoint = 0.5)
```

## Summary

::: {.callout-tip}
#### Key Points

- TODO
:::
